---
title: "03 — Bulk RNA-seq downregulated genes (Tigd4 vs CK8)"
author: "MKim_MTJ"
output:
  html_document:
    toc: true
    toc_float: true
---

> **Reviewer question — Q3.** Which genes are downregulated in MTJ in the bulk RNA-seq CK8 vs Tigd4 contrast (Fig 1F), where the reviewer noted few genes appear down — produce the explicit gene list plus GO terms.
>
> **Purpose.** This report delivers the explicit down-in-MTJ gene list from the Tigd4 vs CK8 DESeq2 contrast and characterises it functionally via GO enrichment, supplying exactly the gene list and GO terms the reviewer requested.

<!--
# Canvas: Prompts/canvases/003_bulk-downregulated.md
Reviewer Q3 (research_plan.md): what are the downregulated genes in Figure 1F
(bulk RNA-seq CK8 vs Tigd4)?

Upstream: RNAseq_Differential_Expression() cached RDS.
Helpers : Scripts/Bin/DE_Reporting.R (classify_DE)
          Scripts/Bin/GO_Enrichment.R (run_GO, go_on_class)
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
suppressPackageStartupMessages({
    library(dplyr)
    library(ggplot2)
    library(ggrepel)
    library(ComplexHeatmap)
})
source.lib("ScanLib.R")
source.lib("Bacon.R")
source.lib("Read_Annotation.R")
source.pro("Differential_Expression.R")
source.pro("DE_Reporting.R")
source.pro("GO_Enrichment.R")

report_name = "03_bulk_downregulated"
outpath     = Path(file.path(.scripts.path, "Results", report_name),
                   format(Sys.Date(), "%y%m%d"))

path_gtf = "/data/local/vfranke/Annotation/mm39/EnsemblGenes/Mus_musculus.GRCm39.113.chr.gtf"
annot    = read_Gene_Annotation(path_gtf)

contrast = "Tigd4_CK8"
```

## Load and classify

```{r}
rna = RNAseq_Differential_Expression()
# $res carries only id/LFC/padj/diff; pull the per-group mean expression from
# $dat to build an MA x-axis (mean normalized expression across the two groups).
rna_res = rna$res %>%
    dplyr::rename(gene_id = id) %>%
    left_join(annot, by = "gene_id") %>%
    left_join(dplyr::select(rna$dat, gene_id, mean.CK8, mean.Tigd4),
              by = "gene_id") %>%
    mutate(baseMean = (mean.CK8 + mean.Tigd4) / 2) %>%
    classify_DE(contrast, lfc_cut = 0.5)

diff_col = paste0("diff.", contrast)
down = rna_res %>%
    filter(.data[[diff_col]] == "Down",
           gene_biotype == "protein_coding")
message("Down genes: ", nrow(down))
```

## Differential expression method

Bulk RNA-seq counts (Salmon → gene level) were tested with **DESeq2** (Wald test) via
`summarize_PigX_RNAseq()`, design `~ Factor`, contrast **Tigd4 vs CK8** (numerator Tigd4,
the MTJ-sorted population). Sample `Tigd4_br2` was excluded upstream on QC. A gene is called
**down in MTJ** when `log2FoldChange.Tigd4_CK8 < -0.5` **and** `padj < 0.05` (Benjamini–
Hochberg), restricted to `gene_biotype == "protein_coding"`. The GO universe is all tested
protein-coding genes. The fold-change cut is **relaxed to 0.5** here (vs the project-wide
`0.75`) because the reviewer noted few genes are called down in MTJ — the looser cut recovers
the weaker-but-consistent down-regulated set.

## Figure 1 — MA plot, down highlighted

*Why this figure:* Labels every down-in-MTJ gene against expression level, making the reviewer's "very few genes are down" claim quantitative and naming each member of the down set the question asks for.

```{r fig1_MA, fig.width=5, fig.height=4}
plot_df = rna_res %>% filter(gene_biotype == "protein_coding") %>%
    mutate(base = .data[["baseMean"]])

g = ggplot(plot_df,
           aes(log10(base + 1),
               .data[[paste0("log2FoldChange.", contrast)]])) +
    geom_point(color = "lightgray", size = 0.5) +
    geom_point(data = dplyr::filter(plot_df, .data[[diff_col]] == "Down"),
               color = "#8CC6E8", size = 0.5) +
    geom_hline(yintercept = 0, linewidth = 0.25) +
    geom_hline(yintercept = -0.5, linetype = 2, linewidth = 0.25) +
    geom_text_repel(data = dplyr::filter(plot_df, .data[[diff_col]] == "Down"),
                    aes(label = gene_name), size = 2.5,
                    min.segment.length = 0, max.overlaps = Inf) +
    labs(x = "log10 baseMean", y = "log2 Tigd4 / CK8") +
    globals$theme
print(g)
ggsave(file.path(outpath, DateNamer("MA_Tigd4_vs_CK8.pdf")),
       g, width = 5, height = 4)
```

*Input:* bulk RNA-seq DE (Tigd4 vs CK8), protein-coding only.
*Method:* baseMean on x vs log2FC on y, down-set highlighted (`classify_DE`, `lfc_cut = 0.5`).
The dashed horizontal line marks the **−0.5** down threshold. **All** down-in-MTJ genes are labelled.
*Shows:* where the down genes lie on the expression spectrum — the reviewer's "very few genes are down" observation.

## Figure 2 — Volcano (Tigd4 vs CK8)

*Why this figure:* puts the reviewer's down set in the full DE context — the 11 down-in-MTJ
genes (light blue) against every tested gene, with the canonical MTJ markers Tigd4/Col22a1
shown for orientation — making plain that the down set is small relative to the up-regulated
MTJ programme.

```{r fig2_volcano, fig.width=5, fig.height=4.5}
volc = rna_res %>% filter(gene_biotype == "protein_coding") %>%
    mutate(lpadj = -log10(.data[[paste0("padj.", contrast)]]),
           lfc   = .data[[paste0("log2FoldChange.", contrast)]])
mark = volc %>% filter(gene_name %in% c("Tigd4", "Col22a1"))

gv = ggplot(volc, aes(lfc, lpadj)) +
    geom_point(color = "lightgray", size = 0.5) +
    geom_point(data = dplyr::filter(volc, .data[[diff_col]] == "Down"),
               color = "#8CC6E8", size = 0.9) +
    geom_point(data = mark, color = "#E88C8C", size = 1.8) +
    geom_vline(xintercept = c(-0.5, 0.5), linetype = 2, linewidth = 0.25) +
    geom_hline(yintercept = -log10(0.05), linetype = 2, linewidth = 0.25) +
    geom_text_repel(data = mark, aes(label = gene_name), size = 3,
                    min.segment.length = 0, max.overlaps = Inf) +
    labs(x = "log2 Tigd4 / CK8", y = "-log10 padj") +
    globals$theme
print(gv)
ggsave(file.path(outpath, DateNamer("volcano_Tigd4_vs_CK8.pdf")), gv, width = 5, height = 4.5)
```

*Input:* bulk RNA-seq DE (Tigd4 vs CK8), protein-coding genes.
*Method:* x = log2FC, y = −log10 padj; the down set (`diff == "Down"`, `lfc_cut = 0.5`) is
highlighted light blue; Tigd4/Col22a1 are drawn in red and text-labelled; dashed lines mark
log2FC = ±0.5 and padj = 0.05.
*Shows:* only a handful of genes are down in MTJ, while the canonical MTJ markers
(Tigd4/Col22a1) sit on the up side — the down programme is small.

## Figure 3 — Volcano: down-regulated translatome (Ribo MTJ/CK8, relaxed −0.5) `(canvas: 022)`

*Why this figure:* the companion to Fig 2 on the **translational** axis — relocated from the
retired report 06. Where Fig 2 shows genes transcriptionally down in MTJ (bulk RNA), this shows
genes whose **translation** is down in MTJ nuclei vs CK8 myofibers (Ribo-seq MTJ/CK8) at the
relaxed −0.5 cut, so the downregulated programme is captured in both modalities.

```{r fig3_ribo_volcano, fig.width=6, fig.height=5}
ribo = Riboseq_Differential_Expression()
ribo_contrast = "MTJ_HA.Rpl22_CK8_HA.Rpl22"
ribo_res = ribo$res %>% dplyr::rename(gene_id = id) %>%
    left_join(annot, by = "gene_id") %>%
    classify_DE(ribo_contrast, lfc_cut = 0.5)

ribo_lfc_v  = paste0("log2FoldChange.", ribo_contrast)
ribo_padj_v = paste0("padj.",           ribo_contrast)
rvolc = ribo_res %>%
    filter(!is.na(.data[[ribo_lfc_v]]), !is.na(.data[[ribo_padj_v]])) %>%
    mutate(log2FC = .data[[ribo_lfc_v]], padj = .data[[ribo_padj_v]],
           Down05 = log2FC < -0.5 & padj < 0.05)

rdown05 = rvolc %>% filter(Down05) %>% arrange(log2FC)
rlab    = head(rdown05, 25)
# superset sanity check: the -0.5 down set must contain the stricter -0.75 set
rdown075 = rvolc %>% filter(log2FC < -0.75 & padj < 0.05)
stopifnot(all(rdown075$gene_id %in% rdown05$gene_id))

gr = ggplot(rvolc, aes(log2FC, -log10(padj))) +
    geom_point(data = dplyr::filter(rvolc, !Down05),
               color = "lightgray", size = 0.6, alpha = 0.6) +
    geom_point(data = dplyr::filter(rvolc, Down05),
               color = "#8CC6E8", size = 0.8, alpha = 0.8) +   # match Fig 2 down colour (canvas 029)
    geom_vline(xintercept = c(-0.5, 0.5), linetype = 2, linewidth = 0.25) +
    geom_hline(yintercept = -log10(0.05), linetype = 2, linewidth = 0.25) +
    geom_text_repel(data = rlab, aes(label = gene_name), size = 2.5,
                    max.overlaps = Inf, min.segment.length = 0, segment.size = 0.2) +
    labs(x = "log2 Ribo MTJ/CK8", y = "-log10 padj") +
    globals$theme
print(gr)
ggsave(file.path(outpath, DateNamer("volcano_ribo_MTJ_CK8_down05.pdf")), gr, width = 6, height = 5)

write.table(rdown05 %>% dplyr::select(gene_id, gene_name, log2FC, padj),
    file = file.path(outpath, DateNamer("table_down05_ribo_MTJ_CK8.tsv")),
    sep = "\t", quote = FALSE, row.names = FALSE)
```

*Input:* cached Ribo-seq DE for the MTJ/CK8 translatome contrast
(`Riboseq_Differential_Expression()`, `MTJ_HA.Rpl22_CK8_HA.Rpl22`), mm39 symbols.
*Method:* down called at the relaxed `log2FC < −0.5 & padj < 0.05`; down set coloured light blue
(`#8CC6E8`, the same down colour as Figure 2) over grey; dashed lines at log2FC ±0.5 and padj 0.05;
top-25 most-negative-log2FC genes labelled;
`stopifnot` asserts the −0.5 set ⊇ the −0.75 set; down table exported.
*Shows:* genes whose translation is most strongly reduced in MTJ nuclei vs CK8 — the
translational counterpart to the transcriptional down set in Fig 2.

## Figure 4 — GO BP on the Ribo-translatome down set `(canvas: 022)`

*Why this figure:* functional characterisation of the translationally-down set (Fig 3),
mirroring the GO done on the bulk-RNA down set — does down-regulated translation in MTJ
concentrate in particular biological processes?

```{r fig4_ribo_go, fig.width=8, fig.height=10}
ribo_down_set = rdown05 %>% filter(gene_biotype == "protein_coding") %>% pull(gene_name) %>% unique()
ribo_universe = ribo_res %>% filter(gene_biotype == "protein_coding") %>% pull(gene_name) %>% unique()
cat(sprintf("GO inputs (canvas 048): %d genes (Ribo-translatome down set) vs universe %d (all tested protein-coding genes in the Ribo MTJ/CK8 DE).\n",
            length(ribo_down_set), length(unique(ribo_universe))))

if(length(ribo_down_set) >= 10){
    go_ribo = run_GO(ribo_down_set, universe = ribo_universe, ont = "BP")
    if(!is.null(go_ribo) && nrow(as.data.frame(go_ribo)) > 0){
        p_ribo = enrichplot::dotplot(go_ribo, showCategory = 20) +
            ggtitle("GO BP — down in MTJ translatome (Ribo MTJ/CK8, -0.5)")
        print(p_ribo)
        ggsave(file.path(outpath, DateNamer("dotplot_GO_BP_ribo_down.pdf")),
               p_ribo, width = 8, height = 10)
        write.table(as.data.frame(go_ribo),
            file = file.path(outpath, DateNamer("table_GO_BP_ribo_down.tsv")),
            sep = "\t", quote = FALSE, row.names = FALSE)
    }
} else {
    cat("Fewer than 10 Ribo-translatome down genes — GO skipped.\n")
}
```

*Input:* the Fig 3 Ribo-translatome down set (protein-coding); universe = all tested protein-coding
genes in the Ribo MTJ/CK8 DE. The exact **gene count and universe size are printed above the figure**
(canvas 048).
*Method:* `run_GO()` (clusterProfiler enrichGO, BP), top-20 terms dotplot; table exported.
*Shows:* the biological processes enriched among translationally-down MTJ genes.

## Down gene list

```{r}
write.table(
    down %>% dplyr::select(gene_id, gene_name, gene_biotype,
                           baseMean,
                           starts_with("log2FoldChange"),
                           starts_with("padj"),
                           starts_with("diff")),
    file = file.path(outpath, DateNamer("table_down_in_MTJ.tsv")),
    sep = "\t", quote = FALSE, row.names = FALSE
)
```

## Figure 5 — Heatmap of downregulated genes (bulk RNA)

*Why this figure:* Confirms the down call is real and consistent across replicates, so the listed genes are genuinely lower in Tigd4 (MTJ) and not artefacts of a single sample.

```{r fig2_down_heatmap, fig.width=5, fig.height=8}
dat = rna$dat %>%
    filter(gene_id %in% down$gene_id) %>%
    dplyr::select(gene_name, contains("sample")) %>%
    as.data.frame() %>%
    magrittr::set_colnames(stringr::str_replace(colnames(.), "sample.", "")) %>%
    magrittr::set_rownames(make.unique(.$gene_name)) %>%
    dplyr::select(-gene_name)

if(nrow(dat) >= 2){
    mat = as.matrix(t(scale(t(log2(dat + 1)))))
    h = Heatmap(mat, name = "z(log2 cnt)",
                show_row_dend = FALSE, show_column_dend = FALSE,
                row_names_gp = grid::gpar(fontsize = 7))
    pdf(file.path(outpath, DateNamer("heatmap_down_in_MTJ.pdf")), width = 5, height = 8)
    draw(h); dev.off()
    draw(h)
} else {
    cat("Fewer than 2 down genes — heatmap skipped.\n")
}
```

*Input:* the down-in-MTJ genes across the bulk RNA samples (CK8 ×4, Tigd4 ×3).
*Method:* row z-scored `log2(count+1)` of the per-sample normalized counts from `rna$dat`.
*Shows:* the down genes should read lower in the Tigd4 (MTJ) columns than the CK8 columns —
a visual confirmation of the DE direction.

## Figures 6-7 — GO enrichment on the bulk-RNA down set

*Why this figure:* The reviewer explicitly asked for GO terms on the downregulated genes; this figure provides the functional characterisation of that down set, turning the bare gene list into the biological-process and cellular-component interpretation requested.

```{r fig2_go_bp}
go_universe_rna = rna_res %>% filter(gene_biotype == "protein_coding") %>% pull(gene_name) %>% unique()
go_down_set_rna = rna_res %>% filter(.data[[diff_col]] == "Down", gene_biotype == "protein_coding") %>%
    pull(gene_name) %>% unique()
cat(sprintf("GO inputs (canvas 048): %d genes (bulk-RNA Down-in-MTJ set) vs universe %d (all tested protein-coding genes in the Tigd4/CK8 DE).\n",
            length(go_down_set_rna), length(go_universe_rna)))

go_bp = go_on_class(rna_res,
                    diff_col = diff_col,
                    class    = "Down",
                    gene_col = "gene_name",
                    universe = go_universe_rna,
                    ont = "BP")

if(!is.null(go_bp) && nrow(as.data.frame(go_bp)) > 0){
    p_bp = enrichplot::dotplot(go_bp, showCategory = 20) +
        ggtitle("GO BP — down in MTJ (Tigd4 vs CK8)")
    print(p_bp)
    ggsave(file.path(outpath, DateNamer("dotplot_GO_BP_down.pdf")),
           p_bp, width = 8, height = 7)
    write.table(as.data.frame(go_bp),
        file = file.path(outpath, DateNamer("table_GO_BP_down.tsv")),
        sep = "\t", quote = FALSE, row.names = FALSE)
}
```

```{r fig3_go_cc}
go_cc = go_on_class(rna_res,
                    diff_col = diff_col,
                    class    = "Down",
                    gene_col = "gene_name",
                    universe = go_universe_rna,   # same set as fig2_go_bp (canvas 048)
                    ont = "CC")

if(!is.null(go_cc) && nrow(as.data.frame(go_cc)) > 0){
    p_cc = enrichplot::dotplot(go_cc, showCategory = 20) +
        ggtitle("GO CC — down in MTJ (Tigd4 vs CK8)")
    print(p_cc)
    ggsave(file.path(outpath, DateNamer("dotplot_GO_CC_down.pdf")),
           p_cc, width = 8, height = 7)
    write.table(as.data.frame(go_cc),
        file = file.path(outpath, DateNamer("table_GO_CC_down.tsv")),
        sep = "\t", quote = FALSE, row.names = FALSE)
}
```

*Input:* gene symbols from the Down class vs universe = all tested protein-coding genes (Tigd4/CK8
DE). The **gene count and universe size are printed above the BP figure** (canvas 048).
*Method:* `clusterProfiler::enrichGO` via `run_GO()` (cached), `OrgDb = org.Mm.eg.db`, BH-adjusted padj < 0.05, minGSSize = 10, maxGSSize = 500.
*Shows:* biological processes (BP) and cellular components (CC) over-represented among the 37 (approx.) down-in-MTJ genes. The reviewer expects this as the main reply artefact.

## Figure 8 — Volcano: all up & down genes labelled (Tigd4 vs CK8) `(canvas: 024)`

*Why this figure:* the companion to the down-focused Figure 2 — it shows the **full
bidirectional** bulk RNA-seq result and labels **every** significant up- and down-regulated gene
(not just the down set / canonical markers), so the complete differentially expressed gene list
is legible directly on the volcano. Because the up set (the MTJ-enriched programme) is large, this
panel is a deliberate completeness artefact; Figure 9 is the readable, unlabelled companion.

```{r fig8_volcano_updown_labeled, fig.width=8, fig.height=7}
volc_ud = rna_res %>% filter(gene_biotype == "protein_coding") %>%
    mutate(lfc   = .data[[paste0("log2FoldChange.", contrast)]],
           lpadj = -log10(.data[[paste0("padj.", contrast)]]),
           dir   = .data[[diff_col]])

n_up   = sum(volc_ud$dir == "Up",   na.rm = TRUE)
n_down = sum(volc_ud$dir == "Down", na.rm = TRUE)
message("Up: ", n_up, "   Down: ", n_down)

sig_ud = volc_ud %>% filter(dir %in% c("Up", "Down"))

g_lab = ggplot(volc_ud, aes(lfc, lpadj)) +
    geom_point(color = "lightgray", size = 0.5) +
    geom_point(data = dplyr::filter(volc_ud, dir == "Up"),   color = "#E88C8C", size = 0.6) +
    geom_point(data = dplyr::filter(volc_ud, dir == "Down"), color = "#8CC6E8", size = 0.6) +
    geom_vline(xintercept = c(-0.5, 0.5), linetype = 2, linewidth = 0.25) +
    geom_hline(yintercept = -log10(0.05), linetype = 2, linewidth = 0.25) +
    geom_text_repel(data = sig_ud, aes(label = gene_name), size = 2,
                    min.segment.length = 0, max.overlaps = Inf, segment.size = 0.2) +
    labs(x = "log2 Tigd4 / CK8", y = "-log10 padj") +
    globals$theme
print(g_lab)
ggsave(file.path(outpath, DateNamer("volcano_Tigd4_CK8_updown_labeled.pdf")),
       g_lab, width = 8, height = 7)
```

*Input:* bulk RNA-seq DE (Tigd4 vs CK8), protein-coding genes; up/down from `classify_DE`,
`lfc_cut = 0.5`, `padj < 0.05`.
*Method:* x = log2FC, y = −log10 padj; up red, down blue, ns light gray; dashed gates at
log2FC = ±0.5 and padj = 0.05. **Every** up ∪ down gene is text-labelled (`max.overlaps = Inf`).
*Shows:* the entire significant up- and down-regulated set named on one plot — `n_up` up vs
`n_down` down (printed above); the up programme dominates, the down set is small (the reviewer's
Q3 observation), seen here against the full DE landscape.

## Figure 9 — Volcano: no gene annotation (Tigd4 vs CK8) `(canvas: 024)`

*Why this figure:* the clean, annotation-free version of Figure 8 — same points, up/down
colouring and significance gates, but no gene-name labels, so the overall shape of the DE
distribution (fold-change spread, up/down balance, significance ceiling) is readable without the
label clutter.

```{r fig9_volcano_bare, fig.width=5, fig.height=4.5}
g_bare = ggplot(volc_ud, aes(lfc, lpadj)) +
    geom_point(color = "lightgray", size = 0.5) +
    geom_point(data = dplyr::filter(volc_ud, dir == "Up"),   color = "#E88C8C", size = 0.6) +
    geom_point(data = dplyr::filter(volc_ud, dir == "Down"), color = "#8CC6E8", size = 0.6) +
    geom_vline(xintercept = c(-0.5, 0.5), linetype = 2, linewidth = 0.25) +
    geom_hline(yintercept = -log10(0.05), linetype = 2, linewidth = 0.25) +
    labs(x = "log2 Tigd4 / CK8", y = "-log10 padj") +
    globals$theme
print(g_bare)
ggsave(file.path(outpath, DateNamer("volcano_Tigd4_CK8_bare.pdf")),
       g_bare, width = 5, height = 4.5)
```

*Input:* identical to Figure 8.
*Method:* same volcano (up red / down blue / ns gray, ±0.5 & padj gates) with **no**
`geom_text_repel` and no highlighted marker points — "without any gene annotation" = no labels,
colours retained.
*Shows:* the unobstructed DE distribution — the readable companion to the fully-labelled Figure 8.

## Up- and down-regulated gene table (bulk RNA) `(canvas: 024)`

```{r table_up_down_RNAseq}
up_down = rna_res %>%
    filter(gene_biotype == "protein_coding", .data[[diff_col]] %in% c("Up", "Down"))

# oracle: the Down rows here must be the same gene set as the legacy down table
stopifnot(setequal(
    up_down %>% filter(.data[[diff_col]] == "Down") %>% pull(gene_id),
    down$gene_id))
# direction sanity: Tigd4/Col22a1 are MTJ markers -> positive log2FC in this Tigd4-numerator contrast
mark_lfc = rna_res %>% filter(gene_name %in% c("Tigd4", "Col22a1")) %>%
    pull(.data[[paste0("log2FoldChange.", contrast)]])
stopifnot(all(mark_lfc > 0))

write.table(
    up_down %>% dplyr::select(gene_id, gene_name, gene_biotype, baseMean,
                              starts_with("log2FoldChange"),
                              starts_with("padj"),
                              starts_with("diff")),
    file = file.path(outpath, DateNamer("table_up_down_RNAseq.tsv")),
    sep = "\t", quote = FALSE, row.names = FALSE)
```

*Input:* bulk RNA-seq DE (Tigd4 vs CK8), protein-coding genes with `diff != "No"` at
`lfc_cut = 0.5`, `padj < 0.05`.
*Method:* one combined table with a `diff.Tigd4_CK8` direction column (Up / Down); the legacy
down-only `table_down_in_MTJ.tsv` is kept unchanged. Two assertions guard it: the Down rows must
equal the legacy down set, and Tigd4/Col22a1 must be up-regulated.
*Shows:* the explicit up and down RNAseq gene lists the reviewer asked for, in a single file.

## Figure 10 — MA plot, unlabeled (Tigd4 vs CK8) `(canvas: 030)`

*Why this figure:* the clean companion to Figure 1 — the same MA (down set highlighted, y=0 and
−0.5 gates) without any gene-name labels, so the overall fold-change-vs-expression distribution
is readable without the label clutter.

```{r fig10_MA_bare, fig.width=5, fig.height=4}
g_ma_bare = ggplot(plot_df,
           aes(log10(base + 1), .data[[paste0("log2FoldChange.", contrast)]])) +
    geom_point(color = "lightgray", size = 0.5) +
    geom_point(data = dplyr::filter(plot_df, .data[[diff_col]] == "Down"),
               color = "#8CC6E8", size = 0.5) +
    geom_hline(yintercept = 0, linewidth = 0.25) +
    geom_hline(yintercept = -0.5, linetype = 2, linewidth = 0.25) +
    labs(x = "log10 baseMean", y = "log2 Tigd4 / CK8") +
    globals$theme
print(g_ma_bare)
ggsave(file.path(outpath, DateNamer("MA_Tigd4_vs_CK8_bare.pdf")),
       g_ma_bare, width = 5, height = 4)
```

*Input:* identical to Figure 1 (bulk RNA-seq DE, protein-coding; down set at `lfc_cut = 0.5`).
*Method:* the Figure-1 MA (x = `log10 baseMean`, y = `log2 Tigd4/CK8`; down set light blue; y=0 and
−0.5 dashed gates) with **no** `geom_text_repel` labels.
*Shows:* the unobstructed MA distribution — the readable companion to the all-down-labelled Figure 1.

## Session info

```{r}
sessionInfo()
```
