---
title: "04 — Exercise translational response classes"
author: "MKim_MTJ"
output:
  html_document:
    toc: true
    toc_float: true
---

> **Reviewer question — Q5.** Which gene classes change translationally with exercise? Are the genes that gain MTJ enrichment specifically after exercise a distinct functional category, or do they share the classification of constitutive MTJ-enriched markers? Characterise both the up and the down responses.
>
> **Purpose.** This report partitions the MTJ translational response into mutually-exclusive classes and isolates the exercise-specific ones — genes translationally up or down after exercise but not in the sedentary state — then runs GO to characterise their function and contrast it with constitutive MTJ enrichment.

<!--
# Canvas: Prompts/canvases/004_exercise-de.md
Reviewer Q5: what gene classes are translationally up / down after exercise?
Are the newly enriched genes sharing MTJ-enriched categories?

Upstream:
  Riboseq_Differential_Expression()           -> sedentary contrast
  Differential_Expression_Exercise()          -> exercise contrast
  Riboseq_Differential_Expression_Combination() -> EXE-vs-untreated swap
  build_exercise_TE()  (Translation_Efficiency.R) -> Fig 7 sedentary-vs-trained TE
Helpers : Scripts/Bin/Exercise_Classification.R (classify_exercise_response)
          Scripts/Bin/GO_Enrichment.R          (run_GO)
          Scripts/Bin/Sample_Sheet.R           (load_sample_sheet)
          Scripts/Bin/Translation_Efficiency.R (build_exercise_TE)  [canvas 020]
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
suppressPackageStartupMessages({
    library(dplyr)
    library(tidyr)
    library(ggplot2)
    library(ggrepel)
    library(ComplexHeatmap)
})
source.lib("ScanLib.R")
source.lib("Bacon.R")
source.lib("Read_Annotation.R")
source.pro("Differential_Expression.R")
source.pro("DE_Reporting.R")
source.pro("GO_Enrichment.R")
source.pro("Exercise_Classification.R")
source.pro("Sample_Sheet.R")             # load_sample_sheet (Fig 7 exercise TE)
source.pro("Translation_Efficiency.R")   # build_exercise_TE (Fig 7 exercise TE)

report_name = "04_exercise_DE"
outpath     = Path(file.path(.scripts.path, "Results", report_name),
                   format(Sys.Date(), "%y%m%d"))

path_gtf = "/data/local/vfranke/Annotation/mm39/EnsemblGenes/Mus_musculus.GRCm39.113.chr.gtf"
annot    = read_Gene_Annotation(path_gtf)
```

## Load cached DE outputs

```{r}
sed = Riboseq_Differential_Expression()
exe = Differential_Expression_Exercise()
cmb = Riboseq_Differential_Expression_Combination()

sed_res = sed$res %>% dplyr::rename(gene_id = id)
exe_res = exe$res %>% dplyr::rename(gene_id = id)
cmb_res = cmb$res %>% dplyr::rename(gene_id = id)
```

## Figure 1 — EXE-vs-WT Ribo-seq volcano (MTJ/CK8 ratio)

*Why this figure:* Establishes the exercise-state MTJ/CK8 translational contrast that defines "changes following exercise" in Q5; without this baseline of significant exercise-state hits there is no set to partition into response classes.

```{r fig1_volcano, fig.width=4, fig.height=4}
contrast_exe = "MTJ_HA_CK8_HA"
plot_df = exe_res %>% left_join(annot, by = "gene_id") %>%
    filter(gene_biotype == "protein_coding") %>%
    classify_DE(contrast_exe) %>%
    mutate(lpval = -log10(.data[[paste0("padj.", contrast_exe)]]))

g1 = ggplot(plot_df,
           aes(.data[[paste0("log2FoldChange.", contrast_exe)]], lpval)) +
    geom_point(color = "lightgray", size = 0.5) +
    geom_point(data = dplyr::filter(plot_df, .data[[paste0("diff.", contrast_exe)]] == "Up"),
               color = "#E88C8C", size = 0.5) +
    geom_point(data = dplyr::filter(plot_df, .data[[paste0("diff.", contrast_exe)]] == "Down"),
               color = "#8CC6E8", size = 0.5) +
    geom_hline(yintercept = -log10(0.05), linetype = 2, linewidth = 0.25) +
    geom_vline(xintercept = c(-0.75, 0.75), linetype = 2, linewidth = 0.25) +
    labs(x = "log2 MTJ-HA / CK8-HA (exercise)", y = "-log10 padj") +
    globals$theme
print(g1)
ggsave(file.path(outpath, DateNamer("volcano_exe_vs_wt.pdf")),
       g1, width = 4, height = 4)
```

*Input:* `Differential_Expression_Exercise()` Ribo-seq contrast `MTJ_HA` vs `CK8_HA`.
*Method:* DESeq2 Wald test, thresholds `|LFC|>0.75`, `padj<0.05`.
*Shows:* MTJ-vs-CK8 enrichment after exercise. Compare with Phase 2 Figure 1 to see which shifts.

## Figure 2 — Sedentary LFC vs exercise LFC

*Why this figure:* Q5 hinges on separating newly exercise-enriched genes from constitutive MTJ markers; plotting sedentary against exercise enrichment is what operationally distinguishes "newly enriched" (off-diagonal) from "already enriched" genes and assigns each to a response class.

```{r fig2_swap_scatter, fig.width=7, fig.height=5}
cls = classify_exercise_response(
    de_sedentary = sed$res %>% dplyr::rename(gene_id = id),
    de_exercise  = exe$res %>% dplyr::rename(gene_id = id),
    de_combined  = cmb$res %>% dplyr::rename(gene_id = id)
) %>% left_join(annot, by = "gene_id")

pc = cls %>% filter(gene_biotype == "protein_coding")

# label set: the four named markers + top-15 exercise-specific genes
# (EXE_only union, ranked by |exercise LFC|)
exe_specific = pc %>%
    filter(response_class %in% c("translationally_up_EXE_only",
                                 "translationally_down_EXE_only")) %>%
    arrange(desc(abs(log2FoldChange_MTJ_HA_CK8_HA))) %>% head(15)
markers = pc %>% filter(gene_name %in% c("Tigd4","Col22a1","Klf2","Klf5"))
lab_df  = dplyr::distinct(dplyr::bind_rows(exe_specific, markers))

# shared response-class colour map (used by Figs 2 & 3, canvas 041): default hue_pal over the
# sorted non-"unchanged" classes (= Fig 2's existing colours), "unchanged" = lightgray
resp_levels = sort(setdiff(unique(cls$response_class), "unchanged"))
class_cols  = setNames(scales::hue_pal()(length(resp_levels)), resp_levels)
class_cols["unchanged"] = "lightgray"

g2 = ggplot(mapping = aes(log2FoldChange_MTJ_HA.Rpl22_CK8_HA.Rpl22,
                          log2FoldChange_MTJ_HA_CK8_HA)) +
    geom_point(data = pc %>% filter(response_class == "unchanged"),
               color = "lightgray", size = 0.5, alpha = 0.5) +
    geom_point(data = pc %>% filter(response_class != "unchanged"),
               aes(color = response_class), size = 0.5, alpha = 0.7) +
    scale_color_manual(values = class_cols) +
    geom_abline(slope = 1, intercept = 0, linetype = 2, linewidth = 0.25) +
    geom_hline(yintercept = 0, linewidth = 0.25) +
    geom_vline(xintercept = 0, linewidth = 0.25) +
    geom_text_repel(data = lab_df, aes(label = gene_name), size = 2.8,
                    min.segment.length = 0, max.overlaps = Inf) +
    labs(x = "log2 MTJ/CK8 sedentary", y = "log2 MTJ/CK8 exercise") +
    globals$theme
print(g2)
ggsave(file.path(outpath, DateNamer("scatter_LFC_sedentary_vs_exercise.pdf")),
       g2, width = 7, height = 5)
```

*Input:* joined sedentary + exercise + combined (swap) DE.
*Method:* `classify_exercise_response()`: eight mutually-exclusive response classes. The
`"unchanged"` class is drawn in light gray; responding classes coloured on top. Labels =
Tigd4/Col22a1/Klf2/Klf5 + the top-15 exercise-specific genes (the two `*_EXE_only` classes,
ranked by `|exercise LFC|`). Panel widened (`fig.width = 7`) so the x-spread is legible.
*Shows:* genes that gain MTJ enrichment specifically under exercise (above diagonal, right side) vs constitutive markers vs losers.

## Figure 3 — Class counts `(canvas: 041)`

*Why this figure:* Quantifies how many genes fall into each response class, so the exercise-specific up/down sets carried into GO are shown to be large enough to be meaningful and to gauge whether the exercise response is dominated by gains, losses, or constitutive enrichment.

```{r fig3_bar, fig.width=7, fig.height=3}
cls_counts = cls %>% dplyr::count(response_class) %>% arrange(-n)
g3 = ggplot(cls_counts, aes(reorder(response_class, n), n, fill = response_class)) +
    geom_col() + coord_flip() +
    scale_fill_manual(values = class_cols) +   # same per-class colours as Figure 2 (canvas 041)
    globals$theme +
    labs(x = NULL, y = "# genes")
print(g3)
ggsave(file.path(outpath, DateNamer("bar_exercise_response_class.pdf")),
       g3, width = 7, height = 3)
```

*Input:* response-class assignment above.
*Method:* `dplyr::count`; bars use the **shared `class_cols` palette** so each class matches its
Figure 2 colour (`unchanged` = lightgray), canvas 041.
*Shows:* size of each translational-response class; the reviewer asks whether `translationally_up_EXE_only` and `translationally_down_EXE_only` differ in function from constitutive MTJ markers.

## Figure 4-5 — GO on EXE-only response classes `(canvas: 026, 038, 042, 048)`

*Why this figure:* This is the direct answer to Q5's functional question — GO on the exercise-specific up and down sets tests whether translationally exercise-induced genes form a distinct functional category and whether the two responses ("each response") enrich different biology.

*GO inputs (canvas 048):* for each class the **number of genes and the universe size are printed above
the dotplots**; universe = protein-coding genes classified for exercise response in `cls`.

```{r go_exe, fig.width=15, fig.height=11}
universe = cls %>% filter(gene_biotype == "protein_coding") %>%
    pull(gene_name) %>% unique()

# per-gene exercise LFC, for ranking a GO term's top-5 genes (canvas 024 -> 026 -> 038)
lfc_map = setNames(cls$log2FoldChange_MTJ_HA_CK8_HA, cls$gene_name)
top5_genes = function(geneID){
    gs = strsplit(geneID, "/")[[1]]
    gs = gs[order(-abs(lfc_map[gs]))]          # strongest exercise responders first
    paste(head(gs, 5), collapse = ", ")
}

go_keep = list()   # accumulate enriched terms per class for the pathway-gene table below

for(cls_name in c("translationally_up_EXE_only","translationally_down_EXE_only")){
    gene_set = cls %>% filter(response_class == cls_name) %>% pull(gene_name) %>% unique()
    if(length(gene_set) < 10){
        message("skipping GO for ", cls_name, ": only ", length(gene_set), " genes")
        next
    }
    cat(sprintf("GO inputs (canvas 048) — %s: %d genes vs universe %d (protein-coding genes classified for exercise response in `cls`).\n",
                cls_name, length(unique(gene_set)), length(unique(universe))))
    go_res = run_GO(gene_symbols = gene_set, universe = universe, ont = "BP")
    if(is.null(go_res) || nrow(as.data.frame(go_res)) == 0) next
    go_keep[[cls_name]] = as.data.frame(go_res)

    p = enrichplot::dotplot(go_res, showCategory = 20) +
        ggtitle(paste0("GO BP — ", cls_name))
    # truncate the p.adjust colourbar tick labels to 3 sig digits, in place (keeps the gradient) — canvas 042
    ci = which(vapply(p$scales$scales,
                      function(s) any(c("colour","color") %in% s$aesthetics), logical(1)))
    if(length(ci)) p$scales$scales[[ci[1]]]$labels = function(x) formatC(x, format = "g", digits = 3)
    # top-5 genes per displayed term (by |exercise LFC|) drawn OUTSIDE the panel; legend below (canvas 038, 042)
    res_df = as.data.frame(go_res)
    res_df$top5 = vapply(res_df$geneID, top5_genes, character(1))
    pd = p$data
    pd$top5 = res_df$top5[match(pd$Description, res_df$Description)]
    xr = max(ggplot2::ggplot_build(p)$data[[1]]$x, na.rm = TRUE)
    p = p + geom_text(data = pd, aes(x = xr * 1.12, y = Description, label = top5),
                      hjust = 0, size = 4.6, inherit.aes = FALSE) +   # font (canvas 045)
        # cap the panel x to the dots so the labels don't stretch the axis; they fall in the margin (canvas 045)
        coord_cartesian(xlim = c(NA, xr * 1.05), clip = "off") +
        theme(plot.margin = unit(c(5.5, 460, 5.5, 5.5), "pt"),
              legend.position = "bottom", legend.box = "horizontal",
              legend.text = element_text(angle = 45, hjust = 1))
    print(p)
    ggsave(file.path(outpath, DateNamer(paste0("dotplot_GO_", cls_name, ".pdf"))),
           p, width = 15, height = 11)
    write.table(as.data.frame(go_res),
        file = file.path(outpath, DateNamer(paste0("table_GO_", cls_name, ".tsv"))),
        sep = "\t", quote = FALSE, row.names = FALSE)
}
```

## Figure 6 — Heatmap of `translationally_up_EXE_only`

*Why this figure:* Resolves the GO summary down to individual exercise-induced genes and their per-sample expression, letting the reviewer verify the exercise-specific up class is a coherent, exercise-driven set rather than a noisy enrichment artefact.

```{r fig6_heatmap, fig.width=6, fig.height=10}
top_up_exe = cls %>%
    filter(response_class == "translationally_up_EXE_only",
           gene_biotype == "protein_coding") %>%
    arrange(-log2FoldChange_MTJ_HA_CK8_HA) %>% head(50)

if(nrow(top_up_exe) >= 5){
    dat = cmb$dat %>%
        filter(gene_id %in% top_up_exe$gene_id) %>%
        dplyr::select(gene_name, contains("sample")) %>%
        as.data.frame() %>%
        magrittr::set_colnames(stringr::str_replace(colnames(.), "sample.", "")) %>%
        magrittr::set_rownames(.$gene_name) %>%
        dplyr::select(-gene_name)
    mat = as.matrix(t(scale(t(log2(dat + 1)))))
    h = Heatmap(mat, show_row_dend = FALSE, show_column_dend = FALSE,
                name = "z(log2 cnt)")
    pdf(file.path(outpath, DateNamer("heatmap_up_EXE_only.pdf")),
        width = 6, height = 10)
    draw(h)
    dev.off()
    draw(h)
}
```

## Figure 7 — Exercise: sedentary vs trained translation efficiency `(canvas: 020)`

*Why this figure:* the other figures partition the MTJ/CK8 enrichment contrast; this one asks the
complementary question on the **translation-efficiency** axis — does training shift MTJ
translation efficiency rather than transcript abundance? The exercise cohort has no matched bulk
RNA, so the MTJ Ribo-seq of each condition is normalized by a matched MTJ snRNA-seq pseudobulk
(dilbaz 2026); plotting sedentary vs trained TE shows the training-induced shift. (Relocated here
from report 02.)

```{r exercise_te}
ss   = load_sample_sheet()
exTE = build_exercise_TE(ss, path_gtf = path_gtf)

# MTJ-label validation (oracle): the dilbaz MTJ nuclei must be Tigd4+/Col22a1+ vs other classes
knitr::kable(exTE$mtj_validation, digits = 3,
    caption = "dilbaz MTJ-label validation — mean normalized Tigd4/Col22a1 in MTJ vs other nuclei. MTJ must be clearly higher.")

ex = exTE$te
knitr::kable(ex %>% dplyr::count(te_class),
    caption = "Genes by training-induced TE class (interaction padj < 0.05 & |Δlog2TE| > 0.75).")
```

```{r exercise_scatter, fig.width=5.5, fig.height=5}
plot_df = ex %>% filter(detectable)
markers_ex = plot_df %>% filter(gene_name %in% c("Tigd4","Col22a1","Klf2","Klf5"))
top_lab    = plot_df %>% filter(significant) %>% arrange(desc(abs(delta_TE))) %>% head(15)
lab_ex     = dplyr::distinct(dplyr::bind_rows(markers_ex, top_lab))

g_ex = ggplot(plot_df, aes(log2TE_sedentary, log2TE_trained)) +
    geom_point(color = "lightgray", size = 0.5) +
    geom_point(data = dplyr::filter(plot_df, te_class == "TE_up_trained"),
               color = "#E88C8C", size = 0.7) +
    geom_point(data = dplyr::filter(plot_df, te_class == "TE_down_trained"),
               color = "#8CC6E8", size = 0.7) +
    geom_point(data = markers_ex, color = "#2CA02C", size = 2.2, shape = 18) +
    geom_abline(slope = 1, intercept = 0, linetype = 2, linewidth = 0.25) +
    geom_text_repel(data = lab_ex, aes(label = gene_name), size = 3,
                    min.segment.length = 0, max.overlaps = 30) +
    labs(x = "log2 TE  sedentary  (Ribo MTJ / dilbaz-MTJ snRNA)",
         y = "log2 TE  trained  (Ribo MTJ / dilbaz-MTJ snRNA)") +
    globals$theme
print(g_ex)
ggsave(file.path(outpath, DateNamer("scatter_TE_sedentary_vs_trained.pdf")),
       g_ex, width = 5.5, height = 5)

write.table(ex %>% dplyr::select(gene_name, log2TE_sedentary, log2TE_trained,
                                 delta_TE, int_padj, te_class, significant),
    file = file.path(outpath, DateNamer("table_exercise_TE.tsv")),
    sep = "\t", quote = FALSE, row.names = FALSE)
```

*Input:* MTJ Ribo-seq — sedentary `riboseq_sedentary` MTJ_HA.Rpl22 (n=3), trained
`riboseq_exercise` MTJ_HA (n=3) — and the dilbaz MTJ snRNA pseudobulk per condition×replicate,
assembled by `build_exercise_TE()`.
*Method:* one joint DESeq2 object over {Ribo,snRNA}×{sedentary,trained}, joint size factors;
per-condition TE = mean(norm Ribo) / mean(norm snRNA pseudobulk). **x = sedentary log2(Ribo
MTJ / dilbaz-MTJ snRNA), y = trained log2(Ribo MTJ / dilbaz-MTJ snRNA).** Red = higher TE when
trained, blue = lower; Tigd4/Col22a1/Klf2/Klf5 marked as green diamonds; labels = those markers
+ the top-15 significant genes by |Δlog2TE|.
*Caveat:* the two Ribo arms are different cohorts (batch ≡ condition) and the snRNA "Trained"
protocol differs from the Ribo exercise protocol — a cross-protocol proxy used for ranking, not
a calibrated FDR.
*Shows:* genes off the diagonal change MTJ translation efficiency with training, separated from
transcript-abundance change because both axes are snRNA-pseudobulk-normalized.

## Figure 8 — Exercise fold-change of top cross-atlas MTJ markers (dilbaz) `(canvas: 044)`

*Why this figure:* takes the genes most robustly MTJ-specific across two independent atlases (the
top-25 of report 05's Kim×Petrany concordance, Fig 5) and asks how each responds to exercise in the
native MTJ — i.e. whether the established MTJ markers are themselves up/down-regulated by training in
the dilbaz snRNA data.

```{r fig8_mtj_marker_exercise_fc, fig.width=6, fig.height=7}
# newest report-05 cross-dataset MTJ table (data-provenance: latest dated, no hardcoded path) — canvas 044
r05_dir = file.path(.scripts.path, "Results", "05_tigd4_specificity_snRNA")
cands   = list.files(r05_dir, pattern = "table_crossdataset_MTJ_kim_vs_petrany\\.tsv$",
                     recursive = TRUE, full.names = TRUE)
stopifnot(length(cands) > 0)
cross_tab = read.delim(sort(cands, decreasing = TRUE)[1], stringsAsFactors = FALSE)

# top-25 MTJ markers by mean cross-atlas enrichment (kim+pet)/2
top25 = cross_tab %>% arrange(desc(mtj_mean)) %>% head(25)

# dilbaz exercise log2FC w.r.t. native (sedentary) MTJ snRNA: log2(trained / sedentary)
fc = top25 %>%
    left_join(ex %>% dplyr::select(gene_name, rna_sed, rna_tr), by = "gene_name") %>%
    mutate(exercise_log2FC = log2((rna_tr + 1) / (rna_sed + 1)))
n_missing = sum(is.na(fc$rna_sed))
message("Fig8 top-25 MTJ markers: ", nrow(fc), " (missing in dilbaz ex: ", n_missing, ")")
fc_plot = fc %>% filter(!is.na(exercise_log2FC))

g8 = ggplot(fc_plot, aes(reorder(gene_name, exercise_log2FC), exercise_log2FC,
                         fill = exercise_log2FC > 0)) +
    geom_col() + coord_flip() +
    scale_fill_manual(values = c(`TRUE` = "#E88C8C", `FALSE` = "#8CC6E8"), guide = "none") +
    geom_hline(yintercept = 0, linewidth = 0.25) +
    labs(x = NULL, y = "exercise log2FC  (dilbaz trained / sedentary MTJ snRNA)",
         title = "Top-25 cross-atlas MTJ markers — exercise response") +
    globals$theme
print(g8)
ggsave(file.path(outpath, DateNamer("bar_mtj_markers_exercise_fc.pdf")), g8, width = 6, height = 7)

write.table(fc %>% dplyr::select(gene_name, kim, pet, mtj_mean, rna_sed, rna_tr, exercise_log2FC),
    file = file.path(outpath, DateNamer("table_mtj_markers_exercise_fc.tsv")),
    sep = "\t", quote = FALSE, row.names = FALSE)
```

*Input:* the top-25 genes by mean MTJ enrichment `(kim+pet)/2` from report 05's newest
`table_crossdataset_MTJ_kim_vs_petrany.tsv` (Kim-17 × Petrany-26 concordance, Fig 5), joined to the
dilbaz MTJ snRNA pseudobulk per-condition expression (`ex$rna_sed`, `ex$rna_tr` from
`build_exercise_TE`).
*Method:* exercise response = `log2((rna_tr + 1)/(rna_sed + 1))` — dilbaz trained-vs-sedentary MTJ
snRNA log2 fold change (native MTJ = sedentary). Horizontal bars per gene, ordered by FC, red = up
with training / blue = down; genes of the 25 not detected in dilbaz are dropped from the bars but
kept (NA) in `table_mtj_markers_exercise_fc.tsv` (count reported).
*Shows:* whether the most reproducible MTJ markers are themselves exercise-regulated in the native
MTJ, and in which direction.

## Pathway-gene expression table `(canvas: 026)`

*Why this table:* makes the GO enrichment (Figs 4-5) traceable to specific genes and their
expression — for every gene in an enriched pathway, its Ribo-seq and dilbaz MTJ snRNA-pseudobulk
expression (sedentary + trained), so the reviewer can see which quantified genes drive each term.

```{r pathway_gene_table}
if(length(go_keep) > 0){
    path_long = do.call(rbind, lapply(names(go_keep), function(cn){
        gdf = go_keep[[cn]]
        do.call(rbind, lapply(seq_len(nrow(gdf)), function(i){
            gs = strsplit(gdf$geneID[i], "/")[[1]]
            data.frame(gene_name = gs, response_class = cn,
                       go_id = gdf$ID[i], go_term = gdf$Description[i],
                       stringsAsFactors = FALSE)
        }))
    }))
    path_genes = path_long %>%
        group_by(gene_name, response_class) %>%
        summarize(n_go_terms = dplyr::n_distinct(go_id),
                  go_terms   = paste(unique(go_term), collapse = "; "),
                  .groups = "drop") %>%
        left_join(dplyr::distinct(cls, gene_name, log2FoldChange_MTJ_HA_CK8_HA),
                  by = "gene_name") %>%
        left_join(ex %>% dplyr::select(gene_name, ribo_sed, ribo_tr, rna_sed, rna_tr),
                  by = "gene_name")
    n_missing = sum(is.na(path_genes$ribo_sed))
    message("Pathway genes: ", nrow(path_genes),
            "  (missing exTE expression: ", n_missing, ")")
    # CSV-only export — the table is no longer rendered inline (canvas 039)
    write.table(path_genes,
        file = file.path(outpath, DateNamer("table_pathway_genes.tsv")),
        sep = "\t", quote = FALSE, row.names = FALSE)
} else {
    cat("No enriched GO terms retained — pathway-gene table skipped.\n")
}
```

*Input:* the enriched GO BP terms per EXE-only class (Figs 4-5, accumulated in `go_keep`), exploded
to member genes; joined to the exercise LFC (`cls`) and the Fig-7 `exTE$te` per-condition expression.
*Method:* per (gene, class) the GO terms are collapsed (`go_terms`, `n_go_terms`); `ribo_sed`/
`ribo_tr` = Ribo-seq MTJ (sedentary/trained), `rna_sed`/`rna_tr` = dilbaz MTJ snRNA pseudobulk.
Genes below the exTE `detectable` filter join as `NA` (count reported, not dropped).
*Output:* the full table is exported to `table_pathway_genes.tsv` only — **not shown inline**
(canvas 039); it has a `response_class` column.

## Export class tables

```{r}
for(cls_name in unique(cls$response_class)){
    sub = cls %>% filter(response_class == cls_name)
    write.table(
        sub %>% dplyr::select(gene_id, gene_name, gene_biotype, response_class,
                              starts_with("log2FoldChange"),
                              starts_with("padj")),
        file = file.path(outpath, DateNamer(paste0("table_", cls_name, ".tsv"))),
        sep = "\t", quote = FALSE, row.names = FALSE)
}
```

## Session info

```{r}
sessionInfo()
```
